$R^*$: A robust MCMC convergence diagnostic with uncertainty using gradient-boosted machines

### 2020-05-06 write_up/ms.tex
updated ar1 figures

\begin{abstract}
  Markov chain Monte Carlo (MCMC) has transformed Bayesian model inference over the past three decades: mainly because of this, Bayesian inference is now a workhorse of applied scientists. Under general conditions, MCMC sampling converges asymptotically to the posterior distribution, but this provides no guarantees about its finite sample performance. The predominant method for monitoring convergence is to run multiple chains and monitor individual chains' characteristics and compare these to the population as a whole: if within-chain and between-chain summaries are comparable, then this is taken to indicate that the chains have converged to a common stationary distribution. Here, we introduce a new method for diagnosing convergence based on whether a machine learning classifier model can successfully discriminate the individual chains. We call this convergence measure $R^*$. In contrast to the predominant $\widehat{R}$, $R^*$ is a single statistic across all parameters that indicates lack of mixing, although individual variables' importance for this metric can also be determined. Additionally, $R^*$ is not based on any single characteristic of the sampling distribution; instead using all the information in the chain, including that given by the joint sampling distribution, which is currently largely overlooked by existing approaches. Since the used machine learning classifier, gradient-boosted regression trees (GBM), provides uncertainty in predictions, as a byproduct, we obtain uncertainty in $R^*$. The method is straightforward to implement, robust to GBM hyperparameter choice, and could be a complementary additional check on MCMC convergence for applied analyses.
\end{abstract}

@@

The most important determination of whether MCMC has worked is whether the sampling distribution has converged to the posterior \citep{brooks2011handbook}. MCMC methods are thus created because of an asymptotic property: that given an infinite number of draws, their sampling distribution approaches the posterior (under general conditions). Although the guarantees are asymptotic, MCMC estimates can have negligible bias with only a relatively small finite number of draws.

@@

To combat this curse of hindsight, running multiple, independent chains, which have been initialised at diverse areas of parameter space is recommended \citep{gelman1992inference}. If the chains appear not to ``mix'' -- a term essentially meaning that it is difficult to resolve an individual chain's path from the mass of paths overlaid on top of one another -- they are yet to converge. This approach makes it less likely that faux-convergence will occur due to chains becoming stuck in an area of parameter space, and running multiple chains is standard practice in applied inference \citep{lambert2018Student}. The predominant approach to quantitatively measuring this mixing is to compare each chain's sampling distribution to that of the population of chains as a whole: specifically, $\widehat{R}$ -- the main convergence statistic used -- compares within-chain variance to that between-chains \citep{gelman1992inference}. If these variances are similar, $\widehat{R}\approx 1$, and chains are deemed to have mixed. Recently, Stan has adopted more advanced variations on the original $\widehat{R}$ formula: for example, splitting individual chains in two to combat poor intra-chain mixing \citep{gelman2013bayesian}; and using ranks of parameter draws rather than the raw values themselves to calculate $\widehat{R}$ \citep{vehtari2019rank}. Additionally, there has been more focus on ensuring that the effective sample size (ESS), a measure of sample quality (see, for example, \citep{lambert2018Student}), is sufficient, and accordingly, new measures of this quantity have been proposed \citep{vehtari2019rank} and adopted \citep{carpenter2017stan}. Collectively, these statistics help alert users of MCMC to issues with sampling (that typically echo issues with the model) meaning that all is not hunky dory.

Here, we introduce $R^*$, a new convergence metric. This statistic is built on the intuition that, if chains are mixed, it should not be possible to discern from a draw's value the chain that generated it. Rephrased, it should not be possible to predict the chain that \textit{caused} a draw. In this vein, we use a machine learning (ML) classifier to measure convergence. Specifically, we train a ML classifier to predict the chain that generated each observation. By evaluating the performance of the classifier on a held-out test set, this provides a new convergence metric. To maximise predictive accuracy, our chosen ML classifier naturally exploits differences in the full joint distributions between the chains, which means it's sensitive to variations across the joint distribution of target model dimensions unlike most existent convergence diagnostics. Our statistic, unlike its $\widehat{R}$ cousins, is scalar valued for multivariate distributions: one model provides a single $R^*$, whereas $\widehat{R}$ has separate values for each univariate marginal distribution. However, the ML classifier we use can straightforwardly be interrogated to determine which parameters were most important for generating predictive accuracy. For our ML classifier, we use gradient-boosted regression trees \citep{friedman2001greedy,greenwell2019package} (``GBM''), since these are known to perform well for the types of tabular data that our problem presents \citep{chollet2018}. For the types of problem we test, $R^*$ calculation is of a speed comparable to some of the newer $\widehat{R}$ measures calculated (typically O(seconds) to calculate), although for models with 10,000s of parameters and many iterations, the time taken is longer. It is also insensitive to GBM's hyperparameters and provides a measure of convergence robust to various Markov chain pathologies. In addition, since GBMs can output predicted class probabilities, we obtain uncertainty measures for $R^*$, which we find provides a useful summary of MCMC convergence. $R^*$ can straightforwardly be incorporated into existing software libraries to provide a complementary convergence metric alongside more established measures.

The structure of this paper is as follows: in \S\ref{sec:method}, we describe in detail the method for calculating $R^*$ and its uncertainty; in \S\ref{sec:results}, we examine the performance of $R^*$ across a range of scenarios introduced in \citep{vehtari2019rank} and elsewhere. Code for reproducing the analyses is provided at \url{https://github.com/ben18785/ml-mcmc-convergence}.

@@

\begin{figure}[!htb]
	\centerline{\includegraphics[width=1.0\textwidth]{../output/unmixed_1.pdf}}
	\caption{\textbf{Chain prediction based on the marginal distribution of a single parameter.} A shows the path of two chains that have mixed (to the right of panel); B shows two chains that have not mixed.}
	\label{fig:marginal}
\end{figure}

@@

The ML classifier we use here is a gradient-boosted regression tree (also known as a type of gradient-boosted machine or GBM, introduced in \cite{friedman2001greedy}), which experience has dictated to be a highly predictive framework for use in tabular data \citep{chollet2018} like ours. Specifically, we use the GBM implementation in \textbf{\textsf{R}}'s ``Caret'' package \citep{kuhn2008building}, which, in turn, uses the ``gbm'' package \citep{greenwell2019package}. The data for each chain has dimensions: $X\in \mathbb{R}^{S}\times \mathbb{R}^{K}$, where $S$ is the number of draws taken (here assumed the same for each chain, but this is not a binding constraint) and $K$ is the number of parameters. We split each chain's draws into randomly divided training and testing tests: here, we use 70\% of draws for training and 30\% for testing. Our GBM model was rapid to execute training then prediction on the testing set (taking $<1$ second on a desktop computer for both these steps for most models we consider in \S\ref{sec:results}), and its predictive performance was insensitive to its hyperparameters. In all the examples explored in \S\ref{sec:results}, the GBM hyperparameter settings we used were: an interaction depth of 3, a shrinkage parameter of 0.1, 10 observations being the minimum required for each node, and that 50 trees would be grown (see \url{https://github.com/ben18785/ml-mcmc-convergence} to replicate this functionality).

We can also get class probabilities, which we leverage to produce an uncertainty distribution for $R^*$. Algorithm \ref{alg:R_star_uncertainty} gives a recipe for generating draws from this distribution, which we now elaborate on in words. For each draw, $s$, in our testing set, GBM outputs a simplex of chain probabilities: $\boldsymbol{p}^{\{s\}}=(p_1^{\{s\}},p_2^{\{s\}},...,p_N^{\{s\}})$, which forms a categorical distribution that can be sampled from to yield a unique chain prediction, $c^{\{s\}}$. By comparing this classification to the true classification, $c^s$, we obtain a binary measure, $a^{\{s\}}=\mathbbm{1}(c^{\{s\}}=c^s)$, of whether this prediction was correct. We repeat this process for each draw in the testing set, generating $\boldsymbol{a}=(a^{\{1\}},a^{\{2\}},...,a^{\{N_\text{test}\}})$, whose average yields a single $R^{*\{i\}}=N \bar{a}$ estimate for iteration $i$. We then iterate this process, for $i=1,2,...,I$, producing a set of $(R^{*\{1\}},R^{*\{2\}},...,R^{*\{I\}})$, which collectively represent a distribution for $R^*$.

\begin{algorithm}[tb]
	\caption{Procedure to generate $I$ draws of $R^*$}
	\label{alg:R_star_uncertainty}
	\begin{algorithmic}
		\STATE Given test data $X_\text{test}$, number of chains $N$, number of iterations $I$, and fitted
		\STATE model, $\text{ML}(x|X_\text{train})\rightarrow(p_1,p_2,...,p_N)$:
		\FOR{$i=1$ to $I$}
		\FOR{$s=1$ to $S_\text{test}$}
		\STATE Obtain test draw, $x^{\{s\}}=X_\text{test}(s)\in \mathbb{R}^K$
		\STATE Predict chain id probabilities, $(p_1^{\{s\}},p_2^{\{s\}},...,p_N^{\{s\}})= \text{ML}(x^{\{s\}}|X_\text{train})$
		\STATE Draw a chain id, $c^{\{s\}} \sim \text{categorical}(p_1^{\{s\}},p_2^{\{s\}},...,p_N^{\{s\}})$
		\STATE Compare with actual id, $c^s$: $a^{\{s\}}=\mathbbm{1}(c^{\{s\}}=c^s)$
		\ENDFOR
		\STATE Calculate predictive accuracy, $\bar{a} = \frac{1}{S_\text{test}} \sum_{s=1}^{S_\text{test}} a^{\{s\}}$
		\STATE Calculate ratio to null model accuracy, $R^{*i} = \bar{a} / (1 / N) = N\bar{a}$
		\ENDFOR
		\RETURN $(R^{*1},R^{*2},...,R^{*I})$
	\end{algorithmic}
\end{algorithm}

@@

\subsection{Heterogeneity in chain variance: autoregressive example}\label{sec:heterogeneity}
We generate four Markov chains, where each samples from an autoregressive order 1 (AR(1)) process of the form,
%
\begin{equation}
X_t = \rho X_{t-1} + \epsilon_t
\end{equation}
%
where $\epsilon_t\stackrel{i.i.d.}{\sim}\mathcal{N}(0, \sigma)$, $\rho=0.3$ and $t=1,2,...,1000$. Three of the chains share the same $\sigma=1$, whereas the other chain has $\sigma=1/3$, so that it has $1/3$ of the (unconditional) variance of the others. In Fig. \ref{fig:ar1}A, we show how a GBM fitted to these data classifies observations according to a draw's value. Unsurprisingly, since the fourth chain has a smaller variance, observations close to zero are likely to be classified as being generated by this chain.

To illustrate the consistency of $R^*$, we perform 1000 replicates where, in each case, we generate four $\{X_t\}$ series as described (i.e. where one chain has a lower variance). We then fit a GBM to a labelled training set. The fitted model is then used to classify draws in an independent test set according to the chain which generated them. For each replicate, we then calculate $R^*$ as described in Algorithm \ref{alg:R_star}. In Fig. \ref{fig:ar1}B, we show that $R^*>1$ for 997/1000 replicates, indicating that the chains have not converged in a vast majority of cases. In Fig. \ref{fig:ar1}C, we show the rank-normalised split-$\widehat{R}$ as calculated in \cite{vehtari2019rank} for each replicate; as for $R^*$, this metric indicates the chains have not converged in all replicates.

@@

To investigate the sensitivity of $R^*$ to the GBM hyperparameters, we perform a sensitivity analysis. In this, we fit GBMs with six different sets of hyperparameters to replicates consisting of the same data as were used to generate Fig. \ref{fig:ar1}. In particular, the GBM hyperparameters we varied were the interaction depth (3, 5 or 7) and the number of trees (50, 100 or 200). The resultant $R^*$ point estimates across 200 replicates for each hyperparameter setting are shown in Fig. \ref{fig:ar1_sensitivity}. These show that hyperparameter choice affects $R^*$, with the highest median values obtained for an interaction depth of 3 with 50 trees (our default across all the examples we consider in this paper). This analysis does suggest, however, that this effect is relatively minor: indeed, across the various hyperparameter sets considered, the percentage of replicates where $R^*<1$ did not strongly vary.

\begin{figure}[!htb]
	\centerline{\includegraphics[width=1.0\textwidth]{../output/ar1_sensitivity.pdf}}
	\caption{\textbf{Autoregressive example sensitivity to GBM hyperparameters.} The rows correspond to different GBM models fit using different hyperparameter sets (as indicated for interaction depth and number of trees and using a shrinkage of 0.1 and a minimum observations per node of 10 in all cases). The horizontal axis shows the $R^*$ values generated using Algorithm \ref{alg:R_star}; points have jitter added and are coloured according to whether they exceed the threshold $R^*=1$, with the percentage below this threshold annotated. The point-ranges shown indicate the 25\%, 50\% and 75\% quantiles for each model. See the R markdown file on the Github repository for code to reproduce this example.}
	\label{fig:ar1_sensitivity}
\end{figure}

\subsection{Diagnosing convergence in joint distributions: multivariate normal models}\label{sec:multivariate_normal}
To illustrate how $R^*$ can diagnose convergence issues in the joint target distribution, we consider a bivariate normal density. In all four chains, we use independent sampling to generate 1000 draws from bivariate normal densities with means of zero; in three of these chains, the covariance matrix is an identity matrix; in one chain, the covariance matrix also has unit diagonal terms but has off-diagonal terms of 0.9, indicating strong covariance between the two dimensions. By construction, all chains target the same marginal distribution in each dimension but the fourth chain has a different joint distribution.

First, we use the code provided in \cite{vehtari2019rank} to calculate rank-normalised $\widehat{R}$ and two different ESS measures, aiming to capture how well certain regions of the posterior have been explored: these are known as bulk ESS and tail ESS. In all cases, the various quantities were calculated based on chains split into halves. For both dimensions, the two ESS measures were above 3700, and $\widehat{R}<1.001$, indicating no issues with convergence.

@@

We next consider a more challenging problem -- a 250-dimensional multivariate normal target where its precision matrix, $\boldsymbol{A}\in\mathbb{R}^{250}\times\mathbb{R}^{250}$, is generated from a Wishart distribution \citep{hoffman2014no}. We assume that the Wishart distribution's degrees of freedom is 250, resulting in a distribution with high correlations between dimensions. We use Stan's NUTS algorithm to sample from this target distribution, and run the algorithm for two different iteration counts (each time across 4 chains): 400 and 10,000 (the latter thinned by a factor of 5). First, we used Stan to sample from the ``centered'' parameterisation of this model, which is of the form,
%
\begin{equation}\label{eq:mvt_normal_250}
\boldsymbol{x}\sim \mathcal{N}(\boldsymbol{0},\boldsymbol{A}^{-1}),
\end{equation}
%
where $\boldsymbol{x}\in\mathbb{R}^{250}$. For each set of draws, we used Algorithm \ref{alg:R_star_uncertainty} to generate an uncertainty distribution for $R^*$, which is shown in Fig. \ref{fig:mvt}A. From the plot for the 400 iteration case, it is clear that convergence has not yet occurred since $R^*>1$ across the bulk of this distribution. Even in the 10,000 iteration case, the $R^*$ distribution remains stubbornly shifted a little rightwards of $R^*=1$ (its mean is 1.03): in this case, $\widehat{R}<1.01$ for 95\% of parameters (Fig. \ref{fig:mvt}B), although 54\% had bulk $\text{ESS}<400$ and 13\% of parameters had tail $\text{ESS}<400$ indicating issues with convergence \citep{vehtari2019rank}.

Rather than run the MCMC sampler for more iterations, we move to a ``non-centered'' parameterisation, which introduces auxillary variables $\boldsymbol{z}\in\mathbb{R}^{250}$ that don't affect $p(\boldsymbol{x})$ but facilitate sampling from it. This model has the form,
%
\begin{align}
\boldsymbol{A}^{-1} = \boldsymbol{L}\boldsymbol{L}^T,\qquad
\boldsymbol{x} = \boldsymbol{L} \boldsymbol{z},\qquad
z_j\sim \mathcal{N}(0, 1), \text{ for } j = 1,2,...,250.
\end{align}
%
where $\boldsymbol{L}$ is the Cholesky decomposition of the covariance matrix, $\boldsymbol{A}^{-1}$. Fig. \ref{fig:mvt}A shows the $R^*$ distribution resultant from 10,000 NUTS iterations in this case: now the distribution is closer to $R^*=1$ (mean $R^*=1.01$). Fig. \ref{fig:mvt}B shows the $\widehat{R}$ values for each $x$ parameter in this model, and, echoing the result for $R^*$, $\widehat{R}<1.01$ in all cases; further, bulk- and tail-$\text{ESS}>400$ for all parameters.

In GBMs, it is possible to calculate variable importance (see, for example, \cite{friedman2001greedy} and \cite{greenwell2019package}), and this allows us to determine which variables were mostly responsible for predictive power. We now compare these with the more established metrics $\widehat{R}$ and ESS as calculated in Stan. For a GBM fitted to the centered model with 10,000 MCMC iterations (thinning by a factor of 5) for each chain, we plot in Fig. \ref{fig:mvt}C variable importance (here high values mean a variable is more important) versus $\widehat{R}$ for all dimensions of the target distribution (including Stan's $lp$ quantity, shown as a triangle). From this plot, there was a positive association between GBM's variable importance and $\widehat{R}$ (Spearman's rank correlation: $\rho=0.33, S=1763958, p<0.01$). In Fig.  \ref{fig:mvt}D, we plot variable importance versus two measures: bulk ESS and tail ESS, which both exhibited a strong non-linear negative association (Spearman's rank correlation: bulk ESS: $\rho=-0.57, S=4142470, p<0.01$; tail ESS: $\rho=-0.56, S=4113709, p<0.01$). Since none of these plots form perfect ``lines'' along which all the plotted points fall, this illustrates that variable importance provides information complementary to $\widehat{R}$ and ESS.

@@

\subsection{Hierarchical model: Eight schools model}\label{sec:eight_shools}
We now examine a simple classic example used to highlight difficulties in performing inference for hierarchical models: referred to as the ``Eight school'' model (see Section 5.5 in \citep{gelman2013bayesian}), which aimed to determine the effects of coaching on SAT scores in eight schools.

>>>>

\begin{abstract}
  Markov chain Monte Carlo (MCMC) has transformed Bayesian model inference over the past three decades: mainly because of this, Bayesian inference is now a workhorse of applied scientists. Under general conditions, MCMC sampling converges asymptotically to the posterior distribution, but this provides no guarantees about its finite sample performance. The predominant method for monitoring convergence is to run multiple chains and monitor individual chains' characteristics and compare these to the population as a whole: if within-chain and between-chain summaries are comparable, then this is taken to indicate that the chains have converged to a common stationary distribution. Here, we introduce a new method for diagnosing convergence based on whether a machine learning classifier model can successfully discriminate the individual chains. We call this convergence measure $R^*$. In contrast to the predominant $\widehat{R}$, $R^*$ is a single statistic across all parameters that indicates lack of mixing, although individual variables' importance for this metric can also be determined. Additionally, $R^*$ is not based on any single characteristic of the sampling distribution; instead using all the information in the chain, including that given by the joint sampling distribution, which is currently largely overlooked by existing approaches. Since our choice of machine learning classifier, a gradient-boosted regression trees model (GBM), provides uncertainty in predictions, as a byproduct, we obtain uncertainty in $R^*$. The method is straightforward to implement, robust to GBM hyperparameter choice, and could be a complementary additional check on MCMC convergence for applied analyses.
\end{abstract}

@@

The most important determination of whether MCMC has worked is whether the sampling distribution has converged to the posterior \citep{brooks2011handbook}. MCMC methods are thus created because of an asymptotic property: that given an infinite number of draws, their sampling distribution approaches the posterior (under general conditions). Although the guarantees are asymptotic, MCMC estimates can have negligible bias with only a relatively small number of draws.

@@

To combat this curse of hindsight, running multiple, independent chains, which have been initialised at diverse areas of parameter space is recommended \citep{gelman1992inference}. If the chains appear not to ``mix'' -- a term essentially meaning that it is difficult to resolve an individual chain's path from the mass of paths overlaid on top of one another -- they are yet to converge. This approach makes it less likely that faux-convergence will occur due to chains becoming stuck in an area of parameter space, and running multiple chains is standard practice in applied inference \citep{lambert2018Student}. The predominant approach to quantitatively measuring this mixing is to compare each chain's sampling distribution to that of the population of chains as a whole: specifically, $\widehat{R}$ -- the main convergence statistic used -- compares within-chain variance to that between-chains \citep{gelman1992inference}. If these variances are similar, $\widehat{R}\approx 1$, and chains are deemed to have mixed. Recently, Stan has adopted more advanced variations on the original $\widehat{R}$ formula: for example, splitting individual chains in two to combat poor intra-chain mixing \citep{gelman2013bayesian}; and using ranks of parameter draws rather than the raw values themselves to calculate $\widehat{R}$ \citep{vehtari2019rank}. Additionally, there has been more focus on ensuring that the effective sample size (ESS), a measure of sample quality (see, for example, \cite{lambert2018Student}), is sufficient, and accordingly, new measures of this quantity have been proposed \citep{vehtari2019rank} and adopted \citep{carpenter2017stan}. Collectively, these statistics help alert users of MCMC to issues with sampling (that typically echo issues with the model) meaning that all is not hunky dory.

Here, we introduce $R^*$, a new convergence metric. This statistic is built on the intuition that, if chains are mixed, it should not be possible to discern from a draw's value the chain that generated it. Rephrased, it should not be possible to predict the chain that \textit{caused} a draw. In this vein, we use a machine learning (ML) classifier to measure convergence. Specifically, we train a ML classifier to predict the chain that generated each observation. By evaluating the performance of the classifier on a held-out test set, this provides a new convergence metric. To maximise predictive accuracy, our chosen ML classifier naturally exploits differences in the full joint distributions between chains, which means it's sensitive to variations across the joint distribution of target model dimensions unlike most existent convergence diagnostics. Our statistic, unlike its $\widehat{R}$ cousins, is scalar valued for multivariate distributions: one model provides a single $R^*$, whereas $\widehat{R}$ has separate values for each univariate marginal distribution. However, the ML classifier we use can straightforwardly be interrogated to determine which parameters were most important for generating predictive accuracy. For our ML classifier, we use gradient-boosted regression trees \citep{friedman2001greedy,greenwell2019package} (``GBM''), since these are known to perform well for the types of tabular data that our problem presents \citep{chollet2018}. For the types of problem we test, $R^*$ calculation is of a speed comparable to some of the newer $\widehat{R}$ measures calculated (typically $\mathcal{O}$(seconds) to calculate), although for models with 10,000s of parameters and many iterations, the time taken is longer. It is also insensitive to GBM's hyperparameters and provides a measure of convergence robust to various Markov chain pathologies. In addition, since GBMs can output predicted class probabilities, we obtain uncertainty measures for $R^*$, which we find provides a useful summary of MCMC convergence. $R^*$ can straightforwardly be incorporated into existing software libraries to provide a complementary convergence metric alongside more established measures.

The structure of this paper is as follows: in \S\ref{sec:method}, we describe in detail the method for calculating $R^*$ and its uncertainty; in \S\ref{sec:results}, we examine the performance of $R^*$ across a range of scenarios introduced in \cite{vehtari2019rank} and elsewhere. Code for reproducing the analyses is provided at \url{https://github.com/ben18785/ml-mcmc-convergence}.

@@

\begin{figure}[!htb]
	\centerline{\includegraphics[width=1.0\textwidth]{../output/unmixed_1.pdf}}
	\caption{\textbf{Chain prediction based on the marginal distribution of a single parameter.} A shows the path of two chains that have mixed (with marginal distribution to the right of panel); B shows two chains that have not mixed.}
	\label{fig:marginal}
\end{figure}

@@

The ML classifier we use here is a gradient-boosted regression tree (also known as a type of gradient-boosted machine or GBM, introduced in \cite{friedman2001greedy}), which experience has dictated to be a highly predictive framework for use in tabular data \citep{chollet2018} like ours. Specifically, we use the GBM implementation in \textbf{\textsf{R}}'s ``Caret'' package \citep{kuhn2008building}, which, in turn, uses the ``gbm'' package \citep{greenwell2019package}. The data for each chain has dimensions: $X\in \mathbb{R}^{S}\times \mathbb{R}^{K}$, where $S$ is the number of draws taken (here assumed the same for each chain, but this is not a binding constraint) and $K$ is the number of parameters. We split each chain's draws into randomly divided training and testing tests: here, we use 70\% of draws for training and 30\% for testing. Our GBM model was rapid to execute training then prediction on the testing set (taking $\mathcal{O}$(seconds) on a desktop computer for both these steps for most models we consider in \S\ref{sec:results}), and its predictive performance was insensitive to its hyperparameters (although we explore this in \S\ref{sec:results}). Unless stated otherwise, in the examples explored in \S\ref{sec:results}, the GBM hyperparameter settings we used were: an interaction depth of 3, a shrinkage parameter of 0.1, 10 observations being the minimum required for each node, and that 50 trees would be grown.

From the GBM fit, predicted chain probabilities can also be obtained, which we leverage to produce an uncertainty distribution for $R^*$. Algorithm \ref{alg:R_star_uncertainty} gives a recipe for generating draws from this distribution, which we now elaborate on in words. For each draw, $s$, in our testing set, GBM outputs a simplex of chain probabilities: $\boldsymbol{p}^{\{s\}}=(p_1^{\{s\}},p_2^{\{s\}},...,p_N^{\{s\}})$, which forms a categorical distribution that can be sampled from to yield a unique chain prediction, $c^{\{s\}}$. By comparing this classification to the true classification, $c^s$, we obtain a binary measure, $a^{\{s\}}=\mathbbm{1}(c^{\{s\}}=c^s)$, of whether this prediction was correct. We repeat this process for each draw in the testing set, generating $\boldsymbol{a}=(a^{\{1\}},a^{\{2\}},...,a^{\{N_\text{test}\}})$, whose average yields a single $R^{*\{i\}}=N \bar{a}$ estimate for iteration $i$. We then iterate this process, for $i=1,2,...,I$, producing a set of $(R^{*\{1\}},R^{*\{2\}},...,R^{*\{I\}})$, which collectively represent a distribution for $R^*$.

\begin{algorithm}[tb]
	\caption{Procedure to generate $I$ draws of $R^*$}
	\label{alg:R_star_uncertainty}
	\begin{algorithmic}
		\STATE Given test data $X_\text{test}$, number of chains $N$, number of iterations $I$, and fitted
		\STATE model, $\text{ML}(x|X_\text{train})\rightarrow(p_1,p_2,...,p_N)$:
		\FOR{$i=1$ to $I$}
		\FOR{$s=1$ to $S_\text{test}$}
		\STATE Obtain test draw, $x^{\{s\}}=X_\text{test}(s)\in \mathbb{R}^K$
		\STATE Predict chain id probabilities, $(p_1^{\{s\}},p_2^{\{s\}},...,p_N^{\{s\}})= \text{ML}(x^{\{s\}}|X_\text{train})$
		\STATE Draw a chain id, $c^{\{s\}} \sim \text{categorical}(p_1^{\{s\}},p_2^{\{s\}},...,p_N^{\{s\}})$
		\STATE Compare with actual id, $c^s$: $a^{\{s\}}=\mathbbm{1}(c^{\{s\}}=c^s)$
		\ENDFOR
		\STATE Calculate predictive accuracy, $\bar{a} = \frac{1}{S_\text{test}} \sum_{s=1}^{S_\text{test}} a^{\{s\}}$
		\STATE Calculate ratio to null model accuracy, $R^{*\{i\}} = \bar{a} / (1 / N) = N\bar{a}$
		\ENDFOR
		\RETURN $(R^{*\{1\}},R^{*\{2\}},...,R^{*\{I\}})$
	\end{algorithmic}
\end{algorithm}

@@

\subsection{Heterogeneity in chain variance: autoregressive example}\label{sec:heterogeneity}
We generate four Markov chains, where each samples from an autoregressive order 1 (AR(1)) process of the form,
%
\begin{equation}
X_t = \rho X_{t-1} + \epsilon_t
\end{equation}
%
where $\epsilon_t\stackrel{i.i.d.}{\sim}\mathcal{N}(0, \sigma)$, $\rho=0.3$ and $t=1,2,...,2000$. Three of the chains share the same $\sigma=1$, whereas the other chain has $\sigma=1/3$, so that it has $1/3$ of the (unconditional) variance of the others. In Fig. \ref{fig:ar1}A, we show how a GBM fitted to these data classifies observations according to a draw's value. Unsurprisingly, since the fourth chain has a smaller variance, observations close to zero are likely to be classified as being generated by this chain.

To illustrate the consistency of $R^*$, we perform 1000 replicates where, in each case, we generate four $\{X_t\}$ series as described (i.e. where one chain has a lower variance). We then fit a GBM to a labelled training set. The fitted model is then used to classify draws in an independent test set according to the chain which generated them. For each replicate, we then calculated $R^*$ as described in Algorithm \ref{alg:R_star}. In Fig. \ref{fig:ar1}B, we show that $R^*>1$ for all replicates, indicating that the chains had not converged in all cases. In Fig. \ref{fig:ar1}C, we show the rank-normalised split-$\widehat{R}$ as calculated in \cite{vehtari2019rank} for each replicate;  as for $R^*$, this metric indicates the chains had not converged in all replicates (as diagnosed by $R^*<1.01$).

@@

To investigate the sensitivity of $R^*$ to the GBM hyperparameters, we performed a sensitivity analysis. In this, we fit GBMs with six different sets of hyperparameters to replicates consisting of the same data used to generate Fig. \ref{fig:ar1}. The hyperparameters we varied were the interaction depth (3, 5 or 7) and the number of trees (50, 100 or 200). The resultant $R^*$ point estimates across 200 replicates for each hyperparameter setting are shown in Fig. \ref{fig:ar1_sensitivity}. These show that hyperparameter choice affects $R^*$, with the highest median values obtained for an interaction depth of 3 with 50 trees (our default across all the examples we consider in this paper). This analysis does suggest, however, that this effect is relatively minor: indeed, across the various hyperparameter sets considered, the percentage of replicates where $R^*<1$ did not strongly vary.

\begin{figure}[!htb]
	\centerline{\includegraphics[width=1.0\textwidth]{../output/ar1_sensitivity.pdf}}
	\caption{\textbf{Autoregressive example sensitivity to GBM hyperparameters.} The rows correspond to GBM models fit using different hyperparameter sets (as indicated for interaction depth and number of trees and using a shrinkage of 0.1 and a minimum observations per node of 10 in all cases). The horizontal axis shows the $R^*$ values generated using Algorithm \ref{alg:R_star}; points have jitter added and are coloured according to whether they exceed the threshold $R^*=1$, with the percentage below this threshold annotated. The point-ranges shown indicate the 25\%, 50\% and 75\% quantiles for each model. See the R markdown file on the Github repository for code to reproduce this example.}
	\label{fig:ar1_sensitivity}
\end{figure}

\subsection{Diagnosing convergence in joint distributions: multivariate normal models}\label{sec:multivariate_normal}
To illustrate how $R^*$ can diagnose convergence issues in the joint target distribution, we consider a bivariate normal density. In all 4 chains, we use independent sampling to generate 1000 draws from bivariate normal densities with means of zero; in three of these chains, the covariance matrix is an identity matrix; in one chain, the covariance matrix also has unit diagonal terms but has off-diagonal terms of 0.9, indicating strong covariance between the two dimensions. By construction, all chains target the same marginal distribution in each dimension but the fourth chain has a different joint distribution.

First, we use the code provided in \cite{vehtari2019rank} to calculate rank-normalised $\widehat{R}$ and two different ESS measures, aiming to capture how well certain regions of the posterior have been explored: these are known as bulk-ESS and tail ESS. In all cases, the various quantities were calculated based on chains split into halves. For both dimensions, the two ESS measures were above 3700, and $\widehat{R}<1.001$, indicating no issues with convergence.

@@

We next consider a more challenging problem -- a 250-dimensional multivariate normal target where its precision matrix, $\boldsymbol{A}\in\mathbb{R}^{250}\times\mathbb{R}^{250}$, is generated from a Wishart distribution \citep{hoffman2014no}. We assume that the Wishart distribution's degrees of freedom is 250, resulting in a distribution with high correlations between dimensions. We use Stan's NUTS algorithm to sample from this target distribution, and run the algorithm for two different iteration counts (each time across 4 chains): 400 and 10,000 (the latter thinned by a factor of 5). First, we used Stan to sample from the ``centered'' parameterisation of this model, which is of the form,
%
\begin{equation}\label{eq:mvt_normal_250}
\boldsymbol{x}\sim \mathcal{N}(\boldsymbol{0},\boldsymbol{A}^{-1}),
\end{equation}
%
where $\boldsymbol{x}\in\mathbb{R}^{250}$. For each set of draws, we used Algorithm \ref{alg:R_star_uncertainty} to generate an uncertainty distribution for $R^*$, which is shown in Fig. \ref{fig:mvt}A. From the plot for the 400 iteration case, it is clear that convergence has not yet occurred since $R^*>1$ across the bulk of this distribution. Even in the 10,000 iteration case, the $R^*$ distribution remains stubbornly shifted a little rightwards of $R^*=1$ (its mean is 1.06): in this case, $\widehat{R}<1.01$ for 95\% of parameters (Fig. \ref{fig:mvt}B), although 54\% had bulk-$\text{ESS}<400$ and 13\% of parameters had tail-$\text{ESS}<400$ indicating issues with convergence \citep{vehtari2019rank}.

Rather than run the MCMC sampler for more iterations, we move to a ``non-centered'' parameterisation, which introduces auxillary variables $\boldsymbol{z}\in\mathbb{R}^{250}$ that don't affect $p(\boldsymbol{x})$ but facilitate sampling from it. This model has the form,
%
\begin{align}
\boldsymbol{A}^{-1} = \boldsymbol{L}\boldsymbol{L}^T,\qquad
\boldsymbol{x} = \boldsymbol{L} \boldsymbol{z},\qquad
z_j\sim \mathcal{N}(0, 1), \text{ for } j = 1,2,...,250.
\end{align}
%
where $\boldsymbol{L}$ is the Cholesky decomposition of the covariance matrix, $\boldsymbol{A}^{-1}$. Fig. \ref{fig:mvt}A shows the $R^*$ distribution resultant from 10,000 NUTS iterations in this case: now the distribution is closer to $R^*=1$ (mean $R^*=1.00$). Fig. \ref{fig:mvt}B shows the $\widehat{R}$ values for each $x$ parameter in this model, and, echoing the result for $R^*$, $\widehat{R}<1.01$ in all cases; further, bulk- and tail-$\text{ESS}>400$ for all parameters.

In GBMs, it is possible to calculate variable importance (see, for example, \cite{friedman2001greedy} and \cite{greenwell2019package}), and this allows us to determine which variables were most informative for predictions. We now compare these with the more established metrics $\widehat{R}$ and ESS. For a GBM fitted to the centered model with 10,000 MCMC iterations (thinning by a factor of 5) for each chain, we plot in Fig. \ref{fig:mvt}C variable importance (here high values mean a variable is more important) versus $\widehat{R}$ for all dimensions of the target distribution (including Stan's $lp$ quantity, shown as a triangle). From this plot, there was a positive association between GBM's variable importance and $\widehat{R}$ (Spearman's rank correlation: $\rho=0.33, S=1763958, p<0.01$). In Fig.  \ref{fig:mvt}D, we plot variable importance versus two measures: bulk-ESS and tail-ESS, which both exhibited a strong non-linear negative association (Spearman's rank correlation: bulk-ESS: $\rho=-0.57, S=4142470, p<0.01$; tail-ESS: $\rho=-0.56, S=4113709, p<0.01$). Since none of these plots form perfect ``lines'' along which all the plotted points fall, this illustrates that variable importance provides information complementary to $\widehat{R}$ and ESS.

@@

\subsection{Hierarchical model: Eight schools model}\label{sec:eight_shools}
We now examine a simple classic example used to highlight difficulties in performing inference for hierarchical models: referred to as the ``Eight schools'' model (see Section 5.5 in \cite{gelman2013bayesian}), which aimed to determine the effects of coaching on SAT scores in eight schools.
